---
title: "Health Inequalities by Gender Identity in All of Us"
author: "Fred Ho"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    fig_width: 12
    fig_height: 8
---

```{r setup, include=FALSE}
# UPDATED CONFIGURATION:
# 1. dev = c("png", "pdf"): Creates both a web-friendly PNG and a vector PDF.
# 2. dpi = 300: Sets high resolution.
# 3. fig.path = "figures/": Saves the files into a "figures" folder.
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  dev = c("png", "pdf"),
  dpi = 300,
  fig.path = "figures/" 
)

# Load libraries
library(arrow)
library(tidyverse)
library(gtsummary)
library(kableExtra)
library(ggplot2)
library(consort)


# 1. Load Data
dat <- read_parquet('../data/cohort_cleaned_20260223.parquet')

# --- G-Formula Data ---
# Load Demographic (Primary) G-Formula
df_gform_means <- read_csv('../results/table_gform_raw_means.csv')
df_gform_contrasts <- read_csv('../results/table_gform_contrasts.csv') %>% 
  mutate(model = "Demographics only")

# Load Sensitivity 1 (SES) G-Formula
df_gform_sens <- read_csv('../results/table_gform_contrasts_sens.csv') %>% 
  mutate(model = "Demographics & SES")

# --- Regression Data ---
# Load Main Regression
df_reg_contrasts <- read_csv('../results/table_reg_contrasts.csv') %>%
  mutate(model = "Demographics only (regression)")

# Load Sensitivity Regression (SES)
df_reg_sens <- read_csv('../results/table_reg_contrasts_sens.csv') %>%
  mutate(model = "Demographics & SES (regression)")

# 2. Define Outcome Groups (Maintaining logical order)
outcomes_numeric <- c("gph_t", "gmh_t", "general_health", "social_roles", 
                      "drinks_per_week",
                      "bmi", "waist_cm", "whtr", "whr", "sbp", "dbp")

outcomes_program_bin <- c('smoking_b', 'vaping_b', "bmi_b", "whtr_b", "whr_b", "hyt_b")

outcomes_ehr_bin <- c("hypertension", "chd", "atrial_fibrillation", "heart_failure", 
                      "stroke_tia", "diabetes", "ckd", "asthma", "copd", 
                      "anxiety_depression", "psychosis_bipolar", "dementia", "epilepsy", 
                      "cancer", "chronic_pain", "ibs", "constipation", "hearing_loss", 
                      "alcohol_disorder", "connective_tissue", "mm")

outcomes_hyt <- c("hyt_uncontrolled", "hyt_undetected")

# 3. Custom Variable Dictionary for Clean Labels
raw_names <- c(outcomes_numeric, outcomes_program_bin, outcomes_ehr_bin, outcomes_hyt)
clean_names_dict <- str_to_title(str_replace_all(raw_names, "_", " "))
names(clean_names_dict) <- raw_names

# Override specific program binary names as requested
clean_names_dict["gph_t"] <- "Global physical health (T-score)"
clean_names_dict["gmh_t"] <- "Global mental health (T-score)"

clean_names_dict["smoking_b"] <- "Current/previous smoking"
clean_names_dict["vaping_b"] <- "Current/previous vaping"
clean_names_dict["bmi_b"] <- "General obesity (BMI)"
clean_names_dict["whtr_b"] <- "Central obesity (WHtR)"
clean_names_dict["whr_b"] <- "Central obesity (WHR)"
clean_names_dict["hyt_b"] <- "Hypertension"

clean_names_dict["waist_cm"] <- "Waist circumference (cm)"
clean_names_dict["whtr"] <- "Waist-height ratio"
clean_names_dict["whr"] <- "Waist hip ratio"

clean_names_dict["bmi"] <- "BMI"
clean_names_dict["sbp"] <- "SBP"
clean_names_dict["dbp"] <- "DBP"

clean_names_dict["chd"] <- "Coronary heart disease"
clean_names_dict["stroke_tia"] <- "Stroke/TIA"

clean_names_dict["anxiety_depression"] <- "Anxiety/depression"
clean_names_dict["psychosis_bipolar"] <- "Psychosis/bipolar disorder"

clean_names_dict["ckd"] <- "Chronic kidney disease"
clean_names_dict["copd"] <- "COPD"
clean_names_dict["ibs"] <- "Inflammatory bowel syndrome"

clean_names_dict["mm"] <- "Multimorbidity"

clean_names_dict["hyt_uncontrolled"] <- "Uncontrolled hypertension"
clean_names_dict["hyt_undetected"] <- "Undiagnosed hypertension"


# Wrap in unique() to prevent the "duplicated level" crash between hyt_b and hypertension
ordered_labels <- unique(unname(clean_names_dict))

# 4. Wrangle Contrasts for Plotting 
df_plot_data <- bind_rows(df_gform_contrasts, df_gform_sens) %>%
  mutate(
    # Apply custom labels and create ordered factor
    outcome_clean = factor(clean_names_dict[outcome], levels = rev(ordered_labels)),
    model = factor(model, levels = c("Demographics only", "Demographics & SES"))
  )

# Wrangle Sens 2 (Regression) for separate plots
df_reg_plot <- bind_rows(df_reg_contrasts, df_reg_sens) %>%
  filter(converged == TRUE) %>%
  mutate(
    outcome_clean = factor(clean_names_dict[outcome], levels = rev(ordered_labels)),
    model = factor(model, levels = c("Demographics only (regression)", "Demographics & SES (regression)"))
  )

# 5. Helper Function for Forest Plots
plot_forest <- function(data, title, x_label) {
  intercept_val <- ifelse(any(data$metric == "ratio"), 1, 0)
  
  ggplot(data, aes(x = estimate, y = outcome_clean, color = model, shape = model)) +
    geom_vline(xintercept = intercept_val, linetype = "dashed", color = "gray50", linewidth = 0.8) +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper), 
                    position = position_dodge(width = 0.6), size = 0.5) +
    facet_grid(reference ~ exposure, scales = "free_y", space = "free") +
    scale_color_manual(values = c("Demographics only" = "#2c3e50", 
                                  "Demographics & SES" = "#e74c3c",
                                  "Demographics only (regression)" = "#27ae60",
                                  "Demographics & SES (regression)" = "#8e44ad")) +
    scale_shape_manual(values = c("Demographics only" = 16, # Circle filled
                                  "Demographics & SES" = 1, # Circle open
                                  "Demographics only (regression)" = 15, # Square filled
                                  "Demographics & SES (regression)" = 0)) + # Square open
    labs(title = title, 
         x = x_label, 
         y = NULL, 
         color = "Model", 
         shape = "Model") +
    theme_bw() +
    theme(
      strip.background = element_rect(fill = "#ecf0f1"),
      strip.text = element_text(face = "bold", size = 10),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 9)
    )
}

# 6. Helper Function for G-Formula Predicted Means Tables
make_means_table <- function(df, target_outcomes, table_caption) {
  df %>%
    filter(converged == TRUE, outcome %in% target_outcomes) %>%
    mutate(
      Outcome = factor(clean_names_dict[outcome], levels = ordered_labels),
      Formatted = sprintf("%.2f (%.2f, %.2f)", estimate, conf.low, conf.high)
    ) %>%
    arrange(Outcome) %>%
    select(Outcome, Gender = gender_c, Formatted) %>%
    pivot_wider(names_from = Gender, values_from = Formatted) %>%
    kbl(booktabs = TRUE, caption = table_caption) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
    scroll_box(height = "75vh", fixed_thead = TRUE)
}

# 7. Helper Function for Table 1 Cell Suppression (<20)
suppress_small_cells <- function(x) {
  n_val <- as.numeric(str_replace_all(str_extract(x, "^[0-9,]+"), ",", ""))
  case_when(
    is.na(n_val) ~ x,         
    n_val < 20 ~ "<20",        
    TRUE ~ x                  
  )
}
```

## 1. Participants flow chart

```{r strobe-flowchart, echo=FALSE, fig.width=6, fig.height=9}

# 1. Define your numbers
n_total <- 633547
n_excl_gender <- 633547 - 623585
n_stage_1 <- 623585
n_excl_intersex <- 623585 - 623166
n_stage_2 <- 623166
n_excl_demo <- 623166 - 588126
n_final <- 588126

# 2. Build the Plot
# The add_box() function chains the boxes together vertically.
# The add_side_box() function automatically places the exclusion to the right.

txt_start <- paste0("Total Registered Participants\n(n = ", format(n_total, big.mark=","), ")")
txt_excl1 <- paste0("Excluded (n = ", format(n_excl_gender, big.mark=","), "):\nMissing gender identity data")
txt_stage1 <- paste0("Participants with\nGender Identity Data\n(n = ", format(n_stage_1, big.mark=","), ")")
txt_excl2 <- paste0("Excluded (n = ", format(n_excl_intersex, big.mark=","), "):\nIntersex / Other sex at birth")
txt_stage2 <- paste0("Participants with\nBinary Sex at Birth\n(n = ", format(n_stage_2, big.mark=","), ")")
txt_excl3 <- paste0("Excluded (n = ", format(n_excl_demo, big.mark=","), "):\nMissing core demographics")
txt_final <- paste0("Final Analytic Cohort\n(n = ", format(n_final, big.mark=","), ")")

plot_consort <- add_box(txt = txt_start) %>%
  add_side_box(txt = txt_excl1) %>% # Automatically branches right
  add_box(txt = txt_stage1) %>%
  add_side_box(txt = txt_excl2) %>%
  add_box(txt = txt_stage2) %>%
  add_side_box(txt = txt_excl3) %>%
  add_box(txt = txt_final)

# 3. Render
plot(plot_consort)
```


## 2. Unadjusted characteristics 

### Sociodemographics

```{r table1a-socio}
dat %>%
  select(gender_c, age_c, race_ethnicity, census_division, 
         education, employment_status, deprivation_quintile) %>%
  tbl_summary(
    by = gender_c,
    missing = "no",
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    label = list(
      age_c ~ "Age (Years)",
      race_ethnicity ~ "Race and Ethnicity",
      census_division ~ "Census Division",
      education ~ "Education Level",
      employment_status ~ "Employment Status",
      deprivation_quintile ~ "Deprivation Quintile"
    )
  ) %>%
  add_overall() %>%
  # ONLY apply suppression to categorical/dichotomous variables
  modify_table_body(~ .x %>% 
    mutate(across(starts_with("stat_"), 
      ~ ifelse(var_type %in% c("categorical", "dichotomous"), suppress_small_cells(.x), .x)
    ))
  ) %>%
  bold_labels() %>%
  as_kable_extra(booktabs = TRUE, caption = "Table 1A: Unadjusted Sociodemographic Characteristics by Gender Identity") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "75vh", fixed_thead = TRUE)
```

## 3. Unadjusted outcomes 

### Numeric outcomes

* T-scores: population mean (SD) = 50 (10); the higher the better
* General health / social roles: 5-point Likert-style response; the higher the better

```{r table1b-numeric}
dat %>%
  select(gender_c, all_of(outcomes_numeric)) %>%
  rename_with(~ clean_names_dict[.x], .cols = -gender_c) %>%
  tbl_summary(
    by = gender_c,
    missing = "no",
    type = list(everything() ~ "continuous"), # Forces Likert variables to be treated as continuous
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  modify_table_body(~ .x %>% 
    mutate(across(starts_with("stat_"), 
      ~ ifelse(var_type %in% c("categorical", "dichotomous"), suppress_small_cells(.x), .x)
    ))
  ) %>%
  bold_labels() %>%
  as_kable_extra(booktabs = TRUE, caption = "Table 1B: Unadjusted Numeric Outcomes by Gender Identity") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "75vh", fixed_thead = TRUE)
```

### Study binary outcomes

```{r table1c-program}
dat %>%
  select(gender_c, all_of(outcomes_program_bin)) %>%
  rename_with(~ clean_names_dict[.x], .cols = -gender_c) %>%
  tbl_summary(
    by = gender_c,
    missing = "no"
  ) %>%
  add_overall() %>%
  modify_table_body(~ .x %>% 
    mutate(across(starts_with("stat_"), 
      ~ ifelse(var_type %in% c("categorical", "dichotomous"), suppress_small_cells(.x), .x)
    ))
  ) %>%
  bold_labels() %>%
  as_kable_extra(booktabs = TRUE, caption = "Table 1C: Unadjusted Program Binary Outcomes by Gender Identity") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "75vh", fixed_thead = TRUE)
```

### EHR derived diagnoses

```{r table1d-ehr}
dat %>%
  select(gender_c, all_of(outcomes_ehr_bin)) %>%
  rename_with(~ clean_names_dict[.x], .cols = -gender_c) %>%
  tbl_summary(
    by = gender_c,
    missing = "no"
  ) %>%
  add_overall() %>%
  modify_table_body(~ .x %>% 
    mutate(across(starts_with("stat_"), 
      ~ ifelse(var_type %in% c("categorical", "dichotomous"), suppress_small_cells(.x), .x)
    ))
  ) %>%
  bold_labels() %>%
  as_kable_extra(booktabs = TRUE, caption = "Table 1D: Unadjusted EHR Derived Diagnoses by Gender Identity") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "75vh", fixed_thead = TRUE)
```


## 4. Adjusted marginal outcomes 

### Numeric Health Outcomes

* T-scores: population mean (SD) = 50 (10); the higher the better
* General health / social roles: 5-point Likert-style response; the higher the better

```{r means-numeric}
make_means_table(df_gform_means, outcomes_numeric, 
                 "Table 2: Marginal means [95% CI] for mumeric outcomes collected in program")
```

> Adjusted for age (P-spline), race/ethnicity, and census division

### Study Binary Outcomes

```{r means-program}
make_means_table(df_gform_means, outcomes_program_bin, 
                 "Table 3: Marginal prevalence [95% CI] for binary outcomes collected in program")
```

> Adjusted for age (P-spline), race/ethnicity, and census division

### EHR-Derived Clinical Diagnoses

```{r means-ehr}
make_means_table(df_gform_means, outcomes_ehr_bin, 
                 "Table 4: Marginal prevalence [95% CI] for EHR-derived diagnoses")
```

> Adjusted for age (P-spline), race/ethnicity, and census division

### Hypertension Care Cascade

* Uncontrolled hypertension defined as measured hypertensive in the program and having an existing hypertension diagnosis in EHR. Compared with people measured normotensive and having an existing hypertension diagnosis in EHR.
* Undiagnosed hypertension defined as measured hypertensive in the program and having no hypertension diagnosis in EHR. Compared with people measured normotensive in program and having no hypertension diagnosis in EHR.

```{r means-hyt}
make_means_table(df_gform_means, outcomes_hyt, 
                 "Table 5: Marginal prevalence [95% CI] for hypertension care cascade")
```

> Adjusted for age (P-spline), race/ethnicity, and census division


## 5. Main findings: comparisons based on g-formula

Primary analysis adjusted for age (P-spline), race/ethnicity, geographic region (census division). 

Sensitivity analysis additionally adjusted for education, area deprivation quintile, and employment status. 

### Mean difference

* T-scores: population mean (SD) = 50 (10); the higher the better
* General health / social roles: 5-point Likert-style response; the higher the better

```{r fig1-numeric, fig.height=10, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_numeric, metric == "difference") %>%
  plot_forest(
    title = "Figure 1: Mean differences in numeric outcomes collected in program",
    x_label = "Mean Difference [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


### Prevalence difference

```{r fig2-program-diff, fig.height=6, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_program_bin, metric == "difference") %>%
  plot_forest(
    title = "Figure 2: Prevalence differences in binary outcomes collected in program",
    x_label = "Prevalence Difference [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


```{r fig3-ehr-diff, fig.height=12, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_ehr_bin, metric == "difference") %>%
  plot_forest(
    title = "Figure 3: Prevalence differences in EHR-derived diagnoses",
    x_label = "Prevalence Difference [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


```{r fig4-hyt-diff, fig.height=4, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_hyt, metric == "difference") %>%
  plot_forest(
    title = "Figure 4: Prevalence differences in hypertension care cascade",
    x_label = "Prevalence Difference [95% CI]"
  )
```

> * **Uncontrolled hypertension** defined as measured hypertensive in the program among people with an existing hypertension diagnosis in EHR.
> * **Undiagnosed hypertension** defined as measured hypertensive in the program among people without an existing hypertension diagnosis in EHR.
> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


### Prevalence ratio

```{r fig5-program-ratio, fig.height=6, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_program_bin, metric == "ratio") %>%
  plot_forest(
    title = "Figure 5: Prevalence ratios in binary outcomes collected in program",
    x_label = "Prevalence Ratio [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


```{r fig6-ehr-ratio, fig.height=12, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_ehr_bin, metric == "ratio") %>%
  plot_forest(
    title = "Figure 6: Prevalence ratios in EHR-derived diagnoses",
    x_label = "Prevalence Ratio [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


```{r fig7-hyt-ratio, fig.height=4, fig.width=12}
df_plot_data %>%
  filter(outcome %in% outcomes_hyt, metric == "ratio") %>%
  plot_forest(
    title = "Figure 7: Prevalence ratios in hypertension care cascade",
    x_label = "Prevalence Ratio [95% CI]"
  )
```

> * **Uncontrolled hypertension** defined as measured hypertensive in the program among people with an existing hypertension diagnosis in EHR.
> * **Undiagnosed hypertension** defined as measured hypertensive in the program among people without an existing hypertension diagnosis in EHR.
> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


## 6. Sensitivity findings: comparisons based on regression only

Note: Regression estimates compute coefficients for numeric variables in linear regression and prevalence ratios (exponent of coefficient) for binary variables in modified Poisson regression.

### Mean difference

* T-scores: population mean (SD) = 50 (10); the higher the better
* General health / social roles: 5-point Likert-style response; the higher the better

```{r fig8-reg-numeric, fig.height=10, fig.width=12}
df_reg_plot %>%
  filter(outcome %in% outcomes_numeric, metric == "difference") %>%
  plot_forest(
    title = "Figure 8: Mean differences for numeric outcomes collected in program (Regression)",
    x_label = "Mean Difference [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


### Prevalence ratio

```{r fig9-reg-program, fig.height=6, fig.width=12}
df_reg_plot %>%
  filter(outcome %in% outcomes_program_bin, metric == "ratio") %>%
  plot_forest(
    title = "Figure 9: Prevalence ratios for binary outcomes collected in programme (Regression)",
    x_label = "Prevalence Ratio [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.


```{r fig10-reg-ehr, fig.height=12, fig.width=12}
df_reg_plot %>%
  filter(outcome %in% outcomes_ehr_bin, metric == "ratio") %>%
  plot_forest(
    title = "Figure 10: Prevalence ratios for EHR-derived diagnoses (Regression)",
    x_label = "Prevalence Ratio [95% CI]"
  )
```

> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.

```{r fig11-reg-hyt, fig.height=4, fig.width=12}
df_reg_plot %>%
  filter(outcome %in% outcomes_hyt, metric == "ratio") %>%
  plot_forest(
    title = "Figure 11: Prevalence ratios for hypertension care cascade (Regression)",
    x_label = "Prevalence Ratio [95% CI]"
  )
```

> * **Uncontrolled hypertension** defined as measured hypertensive in the program among people with an existing hypertension diagnosis in EHR.
> * **Undiagnosed hypertension** defined as measured hypertensive in the program among people without an existing hypertension diagnosis in EHR.
> * **Demographics only model** adjusted for age (P-spline), race/ethnicity, and census division; 
> * **Demographics & SES model** additionally adjusted for education, deprivation quintile, and employment status.